Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-12-23 omrp_lb_fl.tex
All R1 and R2 edits, except for one comment

\begin{abstract}
\noindent Overlapping Batch Means (OBM) has long been used in simulation as a method of reusing data to generate variance estimators with asymptotically lower variance.
In this paper, we apply the OBM method to stochastic programming by formulating a variant of the multiple replications procedure used for assessing solution quality.
We give conditions under which the resulting optimality gap point estimators are strongly consistent, the optimality gap interval estimators are asymptotically valid, and the OBM variance estimators of optimality gap have asymptotically lower variances relative to their nonoverlapping counterparts \citep{Meketon1984,Welch1987}.
We investigate computational efficiency, a combined measure of variance and computation time, providing guidelines on the degree of overlap.  
Numerical experiments on several test problems are presented, examining the small-sample behavior and the empirical computational efficiency of the overlapping batches method in this context.  
\medskip

@@

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \citep{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP.

@@

Consider a covariance stationary stochastic process $y^1, y^2, \dots$ with mean $\mu$ and variance parameter $0<\sigma^2<\infty$.  
The variance parameter is the sum of covariances at all lags, $\sigma^2=\mathrm{Var} \left(y^1 \right)+2\sum_{j=1}^{\infty}\mathrm{Cov} \left(y^1,y^{1+j}\right)$. 
If the variance parameter exists, then  $\sigma^2=\lim_{n\rightarrow \infty}n\mathrm{Var} \left(\ybb\right)$, where $\ybb=\frac{1}{n}\sum_{i=1}^{n} y^{i}$.
The task is to estimate $\mu$ given some realization of the univariate stochastic process $\y = (y^1, y^2, \dots, y^n)$.  
Typically, this involves forming a CI of the form $[L_\alpha(\y), U_\alpha(\y)]$ for a given level of significance $\alpha$ such that $\p{L_\alpha(\y) \leq \mu \leq U_\alpha(\y)} = 1 - \alpha$ or that this probability is approximately $1 - \alpha$ for large enough sample sizes.  
The usual estimator for $\mu$ is the sample mean $\ybb$, which is an unbiased estimator even in the presence of correlated data.  
However, the usual (sample) variance estimator $\frac{1}{n-1} \sum_{i=1}^n (y^i - \ybb)^2$ can be inappropriate to use in the presence of correlated data, resulting in inaccurate interval estimators.  
To overcome this difficulty, various variance estimators have been proposed in the simulation literature; see, e.g., \cite{law_07}.  
We briefly review two of these variance estimators: nonoverlapping and overlapping batch means.

@@

Nonoverlapping batch means takes $n$ observations $y^1, \dots, y^n$ of the stochastic process and splits them into $k$ batches of size $m$, where $k = \frac{n}{m}$ (for simplicity, assume, for now, $n = mk$).  
The sample mean of each batch is then computed as
$
	\yb_j = \frac{1}{m} \sum_{i=1}^{m} y^{m(j-1)+i}, j = 1,2, \dots, k,
$
and the overall sample mean $\ybb = \frac{1}{k} \sum_{j=1}^k \yb_j = \frac{1}{n} \sum_{i=1}^n y^i$ provides a point estimator of $\mu$.  
The variance estimator of the batch means, which is $m/n$ times the sample variance, is calculated as
\begin{equation} \label{eq:var}
	\varhat(\ybb) = \frac{m}{n}\frac{1}{k-1} \sum_{j=1}^k \left( \yb_j - \ybb \right)^2.
\end{equation}
The $(1-\alpha)$-level approximate CI is then formed by $\ybb \pm t_{k-1,\alpha/2} \sqrt{\varhat(\ybb)}$, where $t_{k-1,\alpha/2}$ denotes the $1-\alpha/2$ quantile of the Student's t distribution with $k-1$ degrees of freedom.  
Under appropriate conditions, $n\varhat(\ybb)$ is a consistent estimator of $\sigma^2$.

@@

A $(1-\alpha)$-level approximate CI on the mean can be formed using overlapping batches by $\ybb \pm t_{d_{\gammab}(k-1),\alpha/2}$ $\sqrt{\vartg{\ybb}}$, where the degrees of freedom increase $d_{\gammab}$ is the multiplicative inverse of (\ref{eq:var_reduct_formula}).  
For the maximally overlapping batch means, the degrees of freedom factor is $d_{\gammab} = \frac{3}{2}$.

@@

Taking $n$ observations of $\xit$ and splitting them into $k$ {\it nonoverlapping} batches of size $m$, define
\begin{align} 
	\gb_j  & = \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}), \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         & = \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right], \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress $\xh$ from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to the nonoverlapping $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but on all the observations in that batch,  $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{mj}$ because $\xs_j$ is an optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ batch-means estimators are obtained, the overall mean of these estimators $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The variance estimator is obtained as in (\ref{eq:var}) by $\varhat(\gbb) = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in an approximate $(1-\alpha)$-level one-sided CI, $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varhat(\gbb)} \right]$.

@@

\begin{proof}
	For simplicity of exposition, we drop the subscript $l$ and the $\frac{n_l}{m_l}$ for now and add them in later.  
        Further, we use abbreviations $n_b = \nbl$ to denote the number of batches and $dt = \left(\tfrac{n_l}{m_l} - 1\right) \left(\nbl\right)$ to denote the denominator term in $VG_l$ and $VD_l$. We have
	\begin{align}
		\e{(nVG - nVD)^2} & = \left( \frac{n}{dt} \right)^2 \e{ \left( \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right] \right)^2 } \notag \\
		& \leq \left( \frac{n}{dt} \right)^2 \e{ n_b \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \label{eq:vvar_conv_pf_1}
	\end{align}
	by the Cauchy-Schwartz inequality, where the second vector in the Cauchy-Schwartz inequality is taken as a vector of ones.  
        By examining the batch structure in Figure \ref{fig:overlap_nonint} and the definition of $\gbb$ in (\ref{eq:gbb}), one can see that the terms of the expectation in (\ref{eq:vvar_conv_pf_1}) are not identically distributed.  
        We will separate these into identically distributed ``center'' batches and ``fringe'' batches.  
        For example, in Figure \ref{fig:overlap_nonint}, only batch 4 is a center batch, while the others are fringe batches.  
        Notice in Figure \ref{fig:overlap_nonint}, we see $\left\lceil \gammab^{-1} \right\rceil = 3$ sets of nonoverlapping batches $\{1,4,7\}$, $\{2,5\}$, and $\{3,6\}$, each of which has two fringe batches.  
        In general, there are $\left\lceil \gammab^{-1} \right\rceil$ sets of nonoverlapping batches, each of which will have two fringe batches, and $n_b - 2\left\lceil \gammab^{-1} \right\rceil$ center batches.

	The fringe batches come in identically distributed pairs.  
        The first and last batch are identically distributed, as are the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the same batch structure.  
		  The center batches are identically distributed.  
        Let $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$.
		  Furthermore, let $A_j^c$, $j=\left\lceil \gammab^{-1} \right\rceil+1, \ldots, n_b - \left\lceil \gammab^{-1} \right\rceil$ denote the center batches, and let $A_j^f$, $j=1,\ldots,\left\lceil \gammab^{-1} \right\rceil$ and $A_j^e$, $j=n_b - \left\lceil \gammab^{-1} \right\rceil+1,\ldots,n_b$ denote the fringe batches at the front and end of each set of nonoverlapping batches, respectively.  
		  Continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{dt} \right)^2 n_b \e{ \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{dt} \right)^2 n_b \e{\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left( A_j^f \right)^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left( A_j^c \right)^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left( A_j^e \right)^2 
		} \\
		 = & \left( \frac{n}{dt} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left( A_j^c \right)^2 }
		+ \left( \frac{n}{dt} \right)^2  2 n_b \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left( A_j^f \right)^2 }.
	\end{align*}	
	The last equality follows because each front-fringe batch's $\e{(A_j^f)^2}$, $j=1,\ldots,\left\lceil \gammab^{-1} \right\rceil$  has an identical pair in an end-fringe batch's $\e{(A_j^e)^2}$, $j=n_b - \left\lceil \gammab^{-1} \right\rceil+1,\ldots,n_b$, and all center batches have identical values of $\e{(A_j^c)^2}$, $j=\left\lceil \gammab^{-1} \right\rceil+1, \ldots, n_b - \left\lceil \gammab^{-1} \right\rceil$.  
	As a result, we have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{dt}\right)^2n_b^2 < (2m)^2$.  
        Removing the notation for fringe or center batch for simplicity, each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left( A_j \right)^2 } & = \left( 2m \right)^2 \e{ \left((\gb_j - \gbb) - (\db_j - \dbb)\right)^2 \left((\gb_j - \gbb) + (\db_j - \dbb)\right)^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}} \sqrt{\e{\left((\gb_j - \gbb) + (\db_j - \dbb)\right)^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\left(\frac{n}{m} 4m^2\right)^2 \e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}	& \leq 8 \left( \frac{n}{m} 4m^2\right)^2 \left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

@@

We empirically tested the effectiveness of OMRP on four test problems.
All test problems are two-stage stochastic programs with recourse and were selected to study the behavior of OMRP under a variety of different situations---we will discuss this below.
The characteristics of these problems are given in Table \ref{tb:test_problems}. 
The first test problem is a newsvendor problem with demand distributed uniformly on $\{0, 0.0001, 0.0002, \dots, 10\}$; we denote it as NVD.
The instance of the newsvendor problem can be found in \citep{Bayraksan2006}, except here we used a discrete uniform demand distribution.
The second is a capacity expansion planning problem, denoted CEP1, with random demand vector $\xit$ of dimension 3, with 216 total realizations \citep{higle1996}.
The third is a stochastic knapsack problem denoted 10D, with 10 stochastic parameters and 10 binary decision variables \citep{kleywegt2002sample}.
The final problem is a stochastic vehicle routing problem, denoted DB1, with 46 stochastic parameters and a total of $4.5 \times 10^{25}$ realizations \citep{donohue1995upper}.

@@

The candidate solutions used for NVD and CEP1 and their characteristics can be found in \citep{Bayraksan2006}.
We used the candidate solution (0, 0, 1, 1, 1, 0, 1, 1, 0, 1) for 10D with $\mu_{\xh} = 158.579$ and the candidate solution (11, 14, 8, 11, 7) for DB1 with an estimated $\mu_{\xh} = 1.08$. 
Because the optimal solution of DB1 is unknown, we used a Latin hypercube sample of size 50,000 to estimate $\mu_{\xh}$.
We then used this value of $\mu_{\xh}$ to estimate the coverage probability of the OMRP CIs for DB1.

@@

A summary of results from these experiments is depicted in Figures~\ref{fig:varvar1}--\ref{fig:nv}.
Figures~\ref{fig:varvar1} and \ref{fig:varvar2} show the variance of OMRP variance estimators, normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical relative decreases in variance from \citep{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that lower variances can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
The variance of OMRP variance estimators for DB1 can be less than the asymptotic result established in Theorem \ref{thm:varvar_conv}.
This is because DB1 has not yet been solved to optimality due its size; therefore, it may lack a unique optimal solution. 
Even with the relatively small sample sizes compared to the size of this problem, however, overlapping appears to lower the variances.

\begin{figure}[htb!]
	\centering
	\subfloat[][NVD]{
	   \label{fig:nv_varvar}
	   \includegraphics[width=.5\textwidth]{images/nv_varvar_discrete}}
	\subfloat[][CEP1]{
	   \label{fig:cep1_varvar}
	   \includegraphics[width=.5\textwidth]{images/cep1_varvar}}
	\caption{
		$\var{n VG}$ relative to the nonoverlapping counterpart across various values of $\gamma/m$ ($\gamma/m=1$ denotes the case of nonoverlapping batches) for test problems
		(a) NVD and
		(b) CEP1
	}
\label{fig:varvar1}
\end{figure}

\begin{figure}[htb!]
	\centering
	\subfloat[][10D]{
	   \label{fig:10d_varvar}
	   \includegraphics[width=.5\textwidth]{images/10D_varvar}}
	\subfloat[][DB1]{
	   \label{fig:db1_varvar}
	   \includegraphics[width=.5\textwidth]{images/db1_varvar}}
		\caption{
		$\var{n VG}$ relative to the nonoverlapping counterpart across various values of $\gamma/m$ ($\gamma/m=1$ denotes the nonoverlapping batches) for test problems
		(a) 10D and
		(b) DB1
		}
\label{fig:varvar2}
\end{figure}

>>>>

\begin{abstract}
\noindent Overlapping Batch Means (OBM) has long been used in simulation as a method of reusing data to generate variance estimators with asymptotically lower variance.
In this paper, we apply the OBM method to stochastic programming by formulating a variant of the multiple replications procedure used for assessing solution quality.
We give conditions under which the resulting optimality gap point estimators are strongly consistent, the optimality gap interval estimators are asymptotically valid, and the OBM variance estimators for optimality gap have asymptotically lower variances relative to their nonoverlapping counterparts \citep{Meketon1984,Welch1987}.
We investigate computational efficiency, a combined measure of variance and computation time, providing guidelines on the degree of overlap.  
Numerical experiments on several test problems are presented, examining the small-sample behavior and the empirical computational efficiency of the overlapping batches method in this context.  
\medskip

@@

A natural setting where batch means are used in stochastic optimization (especially in stochastic programming) is for assessing solution quality. 
This typically involves generating point and interval estimators of the optimality gap of a given candidate solution using Monte Carlo simulation. 
Because optimality gap estimators can be difficult to analyze statistically, a common approach is to use nonoverlapping batches of data to generate a number of optimality gap estimators. 
These individual estimators are then used to form point and interval estimators applying the nonoverlapping batch means method.
This is referred to as the Multiple Replications Procedure (MRP) and was initially discussed in \cite{Mak1999}.   
In this paper, we formulate an overlapping batches version of MRP.

@@

Consider a covariance stationary stochastic process $y^1, y^2, \dots$ with mean $\mu$ and variance parameter $0<\sigma^2<\infty$.  
The variance parameter is the sum of covariances at all lags, $\sigma^2=\mathrm{Var} \left(y^1 \right)+2\sum_{j=1}^{\infty}\mathrm{Cov} \left(y^1,y^{1+j}\right)$. 
If the variance parameter exists, then  $\sigma^2=\lim_{n\rightarrow \infty}n\mathrm{Var} \left(\ybb\right)$, where $\ybb=\frac{1}{n}\sum_{i=1}^{n} y^{i}$.
The task is to estimate $\mu$ given some realization of the univariate stochastic process $\y = (y^1, y^2, \dots, y^n)$.  
Typically, this involves forming a CI of the form $[L_\alpha(\y), U_\alpha(\y)]$ for a given level of significance $\alpha$ such that $\p{L_\alpha(\y) \leq \mu \leq U_\alpha(\y)} = 1 - \alpha$ or that this probability is approximately $1 - \alpha$ for large enough sample sizes.  
The usual estimator for $\mu$ is the sample mean $\ybb$, which is an unbiased estimator even in the presence of correlated data.  
However, the usual (sample) variance estimator $\frac{1}{n-1} \sum_{i=1}^n (y^i - \ybb)^2$ can be inappropriate to use in the presence of correlated data, resulting in inaccurate interval estimators.  
To overcome this difficulty, various variance estimators have been proposed in the simulation literature; see, e.g., \cite{law_15}.  
We briefly review two of these variance estimators: nonoverlapping and overlapping batch means.

@@

Nonoverlapping batch means takes $n$ observations $y^1, \dots, y^n$ of the stochastic process and splits them into $k$ batches of size $m$, where $k = \frac{n}{m}$ (for simplicity, assume, for now, $n = mk$).  
In the application under consideration in this article, the observations {\it within} a batch are correlated due to the optimization in each batch; see \S \ref{ssec:mrp} for details. 
The sample mean of each batch is computed as
$
	\yb_j = \frac{1}{m} \sum_{i=1}^{m} y^{m(j-1)+i}, j = 1,2, \dots, k,
$
and the overall sample mean $\ybb = \frac{1}{k} \sum_{j=1}^k \yb_j = \frac{1}{n} \sum_{i=1}^n y^i$ provides a point estimator of $\mu$.  
The variance estimator of the batch means, which is $m/n$ times the sample variance, is calculated as
\begin{equation} \label{eq:var}
	\varhat(\ybb) = \frac{m}{n}\frac{1}{k-1} \sum_{j=1}^k \left( \yb_j - \ybb \right)^2.
\end{equation}
The $(1-\alpha)$-level approximate CI is then formed by $\ybb \pm t_{k-1,\alpha/2} \sqrt{\varhat(\ybb)}$, where $t_{k-1,\alpha/2}$ denotes the $1-\alpha/2$ quantile of the Student's t distribution with $k-1$ degrees of freedom.  
Under appropriate conditions, $n\varhat(\ybb)$ is a consistent estimator of $\sigma^2$ \citep{alexo_enc_06}.

@@

A $(1-\alpha)$-level approximate CI on the mean can be formed using overlapping batches by $\ybb \pm t_{d_{\gammab}(k-1),\alpha/2}$ $\sqrt{\vartg{\ybb}}$, where the degrees of freedom increase $d_{\gammab}$ is the multiplicative inverse of (\ref{eq:var_reduct_formula}).  
For the maximally overlapping batch means, the degrees of freedom increase is $d_{\gammab} = \frac{3}{2}$.

@@

Taking $n$ observations of $\xit$ and splitting them into $k$ {\it nonoverlapping} batches of size $m$, define
\begin{align} 
	\gb_j  & = \frac{1}{m} \sum_{i = 1}^{m} f(\xh, \xit^{m(j-1)+i}) - \min_{\x \in X} \frac{1}{m} \sum_{i=1}^{m} f(\x, \xit^{m(j-1)+i}), \ \ \ \ j = 1,2, \dots, k \label{eq:mrp_gap}\\
         & = \frac{1}{m} \sum_{i = 1}^{m} \left[ f(\xh, \xit^{m(j-1)+i}) -  f(\xs_j, \xit^{m(j-1)+i}) \right], \ \ \ \ j = 1,2, \dots, k, \notag
\end{align}
where $\xs_j$ denotes an optimal solution to (\ref{eq:sto_prog_m}) of batch $j$ given on the right-hand side of (\ref{eq:mrp_gap}).    
Since we are assessing the quality of a given solution $\xh \in \X$, we suppress $\xh$ from the notation $\gb_j$.
The optimality gap estimator $\gb_j$ is similar to the nonoverlapping $\yb_j$ with individual observations $y^{m(j-1)+i} = f(\xh,\xit^{m(j-1)+i}) - f(\xs_j,\xit^{m(j-1)+i})$, for $i=1,\ldots,m$ and $j=1,\ldots, k$.
Notice that in this case, $y^i$ not only depends on observation $i$ of batch $j$, $\xit^{m(j-1)+i}$, but on all the observations in that batch,  $\xit^{m(j-1)+1},\xit^{m(j-1)+2},\ldots,\xit^{mj}$, because $\xs_j$ is an optimal solution to (\ref{eq:sto_prog_m}) given in (\ref{eq:mrp_gap}).   
As before, after $k$ batch-means estimators are obtained, the overall mean of these estimators $\gbb = \frac{1}{k} \sum_{j=1}^k \gb_j$ provides a point estimator of the optimality gap.  
The variance estimator is obtained as in (\ref{eq:var}) by $\varhat(\gbb) = \frac{1}{k} \frac{1}{k-1} \sum_{j=1}^k (\gb_j - \gbb)^2$, which results in an approximate $(1-\alpha)$-level one-sided CI, $\left[0, \gbb + t_{k-1,\alpha} \sqrt{\varhat(\gbb)} \right]$.

@@

\begin{proof}
	For simplicity of exposition, we drop the subscript $l$ and the $\frac{n_l}{m_l}$ for now and add them in later.  
        Further, we use abbreviations $n_b = \nbl$ to denote the number of batches and $\tau = \left(\tfrac{n_l}{m_l} - 1\right) \left(\nbl\right)$ to denote the denominator term in $VG_l$ and $VD_l$. We have
	\begin{align}
		\e{(nVG - nVD)^2} & = \left( \frac{n}{\tau} \right)^2 \e{ \left( \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right] \right)^2 } \notag \\
		& \leq \left( \frac{n}{\tau} \right)^2 \e{ n_b \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \label{eq:vvar_conv_pf_1}
	\end{align}
	by the Cauchy-Schwartz inequality, where the second vector in the Cauchy-Schwartz inequality is taken as a vector of ones.  
        By examining the batch structure in Figure \ref{fig:overlap_nonint} and the definition of $\gbb$ in (\ref{eq:gbb}), one can see that the terms of the expectation in (\ref{eq:vvar_conv_pf_1}) are not identically distributed.  
        We will separate these into identically distributed ``center'' batches and ``fringe'' batches.  
        For example, in Figure \ref{fig:overlap_nonint}, only batch 4 is a center batch, while the others are fringe batches.  
        Notice in Figure \ref{fig:overlap_nonint}, we see $\left\lceil \gammab^{-1} \right\rceil = 3$ sets of nonoverlapping batches $\{1,4,7\}$, $\{2,5\}$, and $\{3,6\}$, each of which has two fringe batches.  
        In general, there are $\left\lceil \gammab^{-1} \right\rceil$ sets of nonoverlapping batches, each of which will have two fringe batches, and $n_b - 2\left\lceil \gammab^{-1} \right\rceil$ center batches.

	The fringe batches come in identically distributed pairs.  
        The first and last batch are identically distributed, as are the second and second-to-last, etc.  
        We know this because we could reverse the order of the i.i.d.\ sample and perform the batching again to get the same batch structure.  
		  The center batches are identically distributed.  
        Let $A_j = (\gb_j - \gbb)^2 - (\db_j - \dbb)^2$.
		  Furthermore, let $A_j^c$, $j=\left\lceil \gammab^{-1} \right\rceil+1, \ldots, n_b - \left\lceil \gammab^{-1} \right\rceil$ denote the center batches, and let $A_j^f$, $j=1,\ldots,\left\lceil \gammab^{-1} \right\rceil$ and $A_j^e$, $j=n_b - \left\lceil \gammab^{-1} \right\rceil+1,\ldots,n_b$ denote the fringe batches at the front and end of each set of nonoverlapping batches, respectively.  
		  Continuing from (\ref{eq:vvar_conv_pf_1}),
	\begin{align*}
		& \left( \frac{n}{\tau} \right)^2 n_b \e{ \sum_{j=1}^{n_b} \left[ (\gb_j - \gbb)^2 - (\db_j - \dbb)^2 \right]^2 } \\
		 =  & \left( \frac{n}{\tau} \right)^2 n_b \e{\sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \left( A_j^f \right)^2 \
		 + \sum_{j=\left\lceil \gammab^{-1} \right\rceil+1}^{n_b-\left\lceil \gammab^{-1} \right\rceil} \left( A_j^c \right)^2 \
		 + \sum_{j=n_b-\left\lceil \gammab^{-1} \right\rceil+1}^{n_b} \left( A_j^e \right)^2 
		} \\
		 = & \left( \frac{n}{\tau} \right)^2 n_b \left(n_b - 2\left\lceil \gammab^{-1} \right\rceil \right) \e{ \left( A_j^c \right)^2 }
		+ \left( \frac{n}{\tau} \right)^2  2 n_b \sum_{j=1}^{\left\lceil \gammab^{-1} \right\rceil} \e{ \left( A_j^f \right)^2 }.
	\end{align*}	
	The last equality follows because each front-fringe batch's $\e{(A_j^f)^2}$, $j=1,\ldots,\left\lceil \gammab^{-1} \right\rceil$  has an identical pair in an end-fringe batch's $\e{(A_j^e)^2}$, $j=n_b - \left\lceil \gammab^{-1} \right\rceil+1,\ldots,n_b$, and all center batches have identical values of $\e{(A_j^c)^2}$, $j=\left\lceil \gammab^{-1} \right\rceil+1, \ldots, n_b - \left\lceil \gammab^{-1} \right\rceil$.  
	As a result, we have $\left\lceil \gammab^{-1} \right\rceil + 1$ expectation terms, each of which has a coefficient that is less than $\left(\frac{n}{\tau}\right)^2n_b^2 < (2m)^2$.  
        Removing the notation for fringe or center batch for simplicity, each expectation can be bounded by
	\begin{align}
		\left( 2m \right)^2 \e{ \left( A_j \right)^2 } & = \left( 2m \right)^2 \e{ \left((\gb_j - \gbb) - (\db_j - \dbb)\right)^2 \left((\gb_j - \gbb) + (\db_j - \dbb)\right)^2 } \notag \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}} \sqrt{\e{\left((\gb_j - \gbb) + (\db_j - \dbb)\right)^4 }} \label{eq:var_reduct_cauchy2} \\
		& \leq \left( 2m \right)^2 \sqrt{\e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}} \sqrt{8 \left(\e{(\gb_j - \gbb)^4} + \e{(\db_j - \dbb)^4} \right)}, \label{eq:var_reduct_holder2}
	\end{align}
	where (\ref{eq:var_reduct_cauchy2}) follows from the Cauchy-Schwartz inequality, and (\ref{eq:var_reduct_holder2}) follows from H\"{o}lder's inequality.  
        Since we can bound both expectation terms in (\ref{eq:var_reduct_holder2}) by A2, we can demonstrate $\sqrt{\frac{n}{m}}(nVG - nVD) \xrightarrow{L^2} 0$ by focusing on only one of the terms.  
        Focusing on the first expectation term in (\ref{eq:var_reduct_holder2}) and returning the $\tfrac{n}{m}$, by H\"{o}lder's inequality,
	\begin{align*}
		\left(\frac{n}{m} 4m^2\right)^2 \e{\left((\gb_j - \gbb) - (\db_j - \dbb)\right)^4}	& \leq 8 \left( \frac{n}{m} 4m^2\right)^2 \left( \e{(\db_j - \gb_j)^4} + \e{(\dbb - \gbb)^4} \right).
	\end{align*}	
	By Lemma \ref{lem:gb_gbb_l4}, both expectation terms converge to zero with exponential rates.  
        With a batching structure $m_l = \lfloor n_l^r \rfloor$ for some $0<r<1$, we see that $\frac{n_l}{m_l} \e{(n_l VG_l - n_l VD_l)^2} \rightarrow 0$.
\end{proof}

@@

We empirically tested the effectiveness of OMRP on four test problems.
All test problems are two-stage stochastic programs with recourse and were selected to study the behavior of OMRP under a variety of different situations---we will discuss this below.
The characteristics of these problems are given in Table \ref{tb:test_problems}. 
The first test problem is a newsvendor problem with demand distributed uniformly on $\{0, 0.0001, 0.0002, \dots, 10\}$; we denote it as NVD.
The instance of the newsvendor problem can be found in \cite{Bayraksan2006}, except here we used a discrete uniform demand distribution.
The second is a capacity expansion planning problem, denoted CEP1, with random demand vector $\xit$ of dimension 3, with 216 total realizations \citep{higle1996}.
The third is a stochastic knapsack problem denoted 10D, with 10 stochastic parameters and 10 binary decision variables \citep{kleywegt2002sample}.
The final problem is a stochastic vehicle routing problem, denoted DB1, with 46 stochastic parameters and a total of $4.5 \times 10^{25}$ realizations \citep{donohue1995upper}.

@@

The candidate solutions used for NVD and CEP1 and their characteristics can be found in \cite{Bayraksan2006}.
We used the candidate solution (0, 0, 1, 1, 1, 0, 1, 1, 0, 1) for 10D with $\mu_{\xh} = 158.579$ and the candidate solution (11, 14, 8, 11, 7) for DB1 with an estimated $\mu_{\xh} = 1.08$. 
Because the optimal solution of DB1 is unknown, we used a Latin hypercube sample of size 50,000 to estimate $\mu_{\xh}$.
We then used this value of $\mu_{\xh}$ to estimate the coverage probability of the OMRP CIs for DB1.

@@

A summary of results from these experiments is depicted in Figures~\ref{fig:varvar1} and \ref{fig:nv}.
Figure~\ref{fig:varvar1} shows the variance of OMRP variance estimators, normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical relative decreases in variance from \cite{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1, 10D, and DB1. 
These experiments suggest that lower variances can be achieved in stochastic programming by overlapping the batches, even with small sample sizes.
The variance of OMRP variance estimators for DB1 can be less than the asymptotic result established in Theorem \ref{thm:varvar_conv}.
This is because DB1 has not yet been solved to optimality due its size; therefore, it may lack a unique optimal solution. 
Even with the relatively small sample sizes compared to the size of this problem, however, overlapping appears to lower the variances.

\begin{figure}[htb!]
	\centering
	\subfloat[][NVD]{
	   \label{fig:nv_varvar}
	   \includegraphics[width=.5\textwidth]{images/nv_varvar_discrete}}
	\subfloat[][CEP1]{
	   \label{fig:cep1_varvar}
	   \includegraphics[width=.5\textwidth]{images/cep1_varvar}}
	\\
	\subfloat[][10D]{
	   \label{fig:10d_varvar}
	   \includegraphics[width=.5\textwidth]{images/10D_varvar}}
	\subfloat[][DB1]{
	   \label{fig:db1_varvar}
	   \includegraphics[width=.5\textwidth]{images/db1_varvar}}
	\caption{
		$\var{n VG}$ relative to the nonoverlapping counterpart across various values of $\gamma/m$ ($\gamma/m=1$ denotes the case of nonoverlapping batches) for test problems
		(a) NVD,
		(b) CEP1,
		(c) 10D, and
		(d) DB1
	}
\label{fig:varvar1}
\end{figure}
